dads manual

### 2020-12-04 inst/gitbook/_book/dads_manual.tex
Updated manual (running)

\title{dads manual}
\author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}), Alex Slavenko (email) and others}
\date{2020-11-13}

@@

{
\setcounter{tocdepth}{1}
\tableofcontents
}
\hypertarget{dads}{%
\chapter{\texorpdfstring{\texttt{dads}}{dads}}\label{dads}}

Disparity and diversity simulations.

@@

\begin{verbatim}
##  ---- dads modifiers object ---- 
## No modifiers applied to the branch length and speciation processes (default).
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\CommentTok{## Setting the simulation parameters}
\NormalTok{extinction_}\DecValTok{02}\NormalTok{ <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \FloatTok{0.4}\NormalTok{)}
\NormalTok{living_}\DecValTok{20}\NormalTok{     <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{20}\NormalTok{)}
\NormalTok{BM_trait      <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{4242}\NormalTok{)}
\CommentTok{## Simulate disparity and diversity}
\NormalTok{default_data <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction_}\DecValTok{02}\NormalTok{,}
                     \DataTypeTok{stop.rule =}\NormalTok{ living_}\DecValTok{20}\NormalTok{,}
                     \DataTypeTok{traits    =}\NormalTok{ BM_trait,}
                     \DataTypeTok{modifiers =}\NormalTok{ default_modifiers)}
\NormalTok{default_data}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads object ---- 
## Birth death process with modifiers:
## No modifiers applied to the branch length and speciation processes (default).
## 
## Simulated diversity data (x$tree):
## 
## Phylogenetic tree with 45 tips and 44 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
## 
## Simulated disparity data (x$data):
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Triggering a modification only if the ancestor trait is negative}
\NormalTok{negative.ancestor <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, parent.lineage) \{}
    \KeywordTok{return}\NormalTok{(}\KeywordTok{all}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, parent.lineage) }\OperatorTok{<}\StringTok{ }\DecValTok{0}\NormalTok{))}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

@@

\begin{verbatim}
##  ---- dads modifiers object ---- 
## Default branch length process.
## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (staying.alive).
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{4242}\NormalTok{)}
\CommentTok{## Simulate disparity and diversity}
\NormalTok{biased_data <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction_}\DecValTok{02}\NormalTok{,}
                    \DataTypeTok{stop.rule =}\NormalTok{ living_}\DecValTok{20}\NormalTok{,}
                    \DataTypeTok{traits    =}\NormalTok{ BM_trait,}
                    \DataTypeTok{modifiers =}\NormalTok{ negatives_extinct)}
\NormalTok{biased_data}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads object ---- 
## Birth death process with modifiers:
## Default branch length process.
## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (staying.alive).
## 
## Simulated diversity data (x$tree):
## 
## Phylogenetic tree with 38 tips and 37 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
## 
## Simulated disparity data (x$data):
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

@@

\begin{verbatim}
## $A
## $A$process
## function (x0, edge.length = 1, Sigma = diag(length(x0)), ...) 
## {
##     return(t(MASS::mvrnorm(n = 1, mu = x0, Sigma = Sigma * edge.length, 
##         ...)))
## }
## <bytecode: 0x7fd7168d5b80>
## <environment: namespace:dads>
## 
## $A$start
## [1] 0
## 
## $A$trait_id
## [1] 1
\end{verbatim}

@@

\begin{quote}
TODO: just ideas so far
\end{quote}

@@

A branch length function that is constant

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{constant.brlen <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{bd.params =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{n.taxa =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{parent.lineage =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{trait.values =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{modify.fun =} \OtherTok{NULL}\NormalTok{) \{}
    \KeywordTok{return}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{\}}

\NormalTok{constant_brlen <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \FloatTok{0.25}\NormalTok{),}
                     \DataTypeTok{stop.rule =} \KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{20}\NormalTok{),}
                     \DataTypeTok{traits    =} \KeywordTok{make.traits}\NormalTok{(),}
                     \DataTypeTok{modifiers =} \KeywordTok{make.modifiers}\NormalTok{(}
                        \DataTypeTok{branch.length =}\NormalTok{ constant.brlen))}

\KeywordTok{plot}\NormalTok{(constant_brlen, }\DataTypeTok{main =} \StringTok{"constant branch length"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{random.brlen <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(}\DataTypeTok{bd.params =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{n.taxa =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{parent.lineage =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{trait.values =} \OtherTok{NULL}\NormalTok{,}
                         \DataTypeTok{modify.fun =} \OtherTok{NULL}\NormalTok{) \{}
    \KeywordTok{return}\NormalTok{(}\KeywordTok{runif}\NormalTok{(}\DecValTok{1}\NormalTok{))}
\NormalTok{\}}

>>>>

\title{dads manual}
\author{Thomas Guillerme (\href{mailto:guillert@tcd.ie}{\nolinkurl{guillert@tcd.ie}}), Alex Slavenko (email) and others}
\date{2020-12-04}

@@

{
\setcounter{tocdepth}{1}
\tableofcontents
}
\hypertarget{dads-disparity-and-diversity-simulations.}{%
\chapter{\texorpdfstring{\texttt{dads}: disparity and diversity simulations.}{dads: disparity and diversity simulations.}}\label{dads-disparity-and-diversity-simulations.}}

@@

\begin{verbatim}
##  ---- dads modifiers object ---- 
## No modifiers applied to the branch length, selection and speciation processes (default).
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Setting the simulation parameters}
\NormalTok{extinction_}\DecValTok{02}\NormalTok{ <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \FloatTok{0.2}\NormalTok{)}
\NormalTok{living_}\DecValTok{20}\NormalTok{     <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{20}\NormalTok{)}
\NormalTok{BM_trait      <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\CommentTok{## Simulate disparity and diversity}
\NormalTok{default_data <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction_}\DecValTok{02}\NormalTok{,}
                     \DataTypeTok{stop.rule =}\NormalTok{ living_}\DecValTok{20}\NormalTok{,}
                     \DataTypeTok{traits    =}\NormalTok{ BM_trait,}
                     \DataTypeTok{modifiers =}\NormalTok{ default_modifiers)}
\NormalTok{default_data}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads object ---- 
## Birth death process with modifiers:
## No modifiers applied to the branch length, selection and speciation processes (default).
## 
## Simulated diversity data (x$tree):
## 
## Phylogenetic tree with 24 tips and 23 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
## 
## Simulated disparity data (x$data):
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Triggering a modification only if the ancestor trait is negative}
\NormalTok{negative.ancestor <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, lineage) \{}
    \KeywordTok{return}\NormalTok{(}\KeywordTok{all}\NormalTok{(}\KeywordTok{parent.traits}\NormalTok{(trait.values, lineage) }\OperatorTok{<}\StringTok{ }\DecValTok{0}\NormalTok{))}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

@@

\begin{verbatim}
##  ---- dads modifiers object ---- 
## Default branch length process.
## Default selection process.
## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (staying.alive).
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\CommentTok{## Simulate disparity and diversity}
\NormalTok{biased_data <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ extinction_}\DecValTok{02}\NormalTok{,}
                    \DataTypeTok{stop.rule =}\NormalTok{ living_}\DecValTok{20}\NormalTok{,}
                    \DataTypeTok{traits    =}\NormalTok{ BM_trait,}
                    \DataTypeTok{modifiers =}\NormalTok{ negatives_extinct)}
\NormalTok{biased_data}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads object ---- 
## Birth death process with modifiers:
## Default branch length process.
## Default selection process.
## Speciation process is set to speciation with a condition (negative.ancestor) and a modifier (staying.alive).
## 
## Simulated diversity data (x$tree):
## 
## Phylogenetic tree with 22 tips and 21 internal nodes.
## 
## Tip labels:
##   t1, t2, t3, t4, t5, t6, ...
## Node labels:
##   n1, n2, n3, n4, n5, n6, ...
## 
## Rooted; includes branch lengths.
## 
## Simulated disparity data (x$data):
## 1 trait for 1 process (A) with one starting value (0).
\end{verbatim}

@@

\begin{verbatim}
## $A
## $A$process
## function (x0, edge.length = 1, Sigma = diag(length(x0)), ...) 
## {
##     return(t(MASS::mvrnorm(n = 1, mu = x0, Sigma = Sigma * edge.length, 
##         ...)))
## }
## <bytecode: 0x7f7eeee57f58>
## <environment: namespace:dads>
## 
## $A$start
## [1] 0
## 
## $A$trait_id
## [1] 1
\end{verbatim}

@@

\texttt{"modifiers"} have a similar structure than \texttt{"traits"} where you can design an object with increasing complexity, starting with the simplest modifiers that doesn't modify anything (using the default arguments):

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Making a default modifier (no modification)}
\NormalTok{my_default_modifiers <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{()}
\NormalTok{my_default_modifiers}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  ---- dads modifiers object ---- 
## No modifiers applied to the branch length, selection and speciation processes (default).
\end{verbatim}

Similarly to \texttt{"traits"} objects, \texttt{"modifiers"} are also printed by default using \texttt{print.dads}.
You can see details about what's actually in the object using \texttt{print.dads(my\_default\_modifiers,\ all\ =\ TRUE)}.
However, contrary to \texttt{"traits"}, you cannot plot \texttt{"modifiers"}.

@@

The first argument in \texttt{"modifiers"} is the branch length function (\texttt{branch.length}) this is the function that will be executed in \texttt{dads} to generate branch length.
Note that in the \texttt{dads} algorithm, branch length is not generated \emph{directly} but actually results of the waiting time.
In other words, the \texttt{branch.length} function just affects waiting time for all taxa present at any time in the simulation.
These taxa can then either go extinct (stopping the ``growth'' of it's branch length) or survive (continuing the ``growth'').

By default, branch length (or waiting or growth) is a randomly drawn number from an exponential distribution with the rate of the number of taxa multiplied by the speciation and extinction rate: \(n \times (\lambda + \mu)\) (where \(n\) is the number of taxa currently present in the simulation, \(\lambda\) and \(\mu\) are respectively the speciation and extinction rates).
This default function is simply called \texttt{branch.length} in \texttt{dads} and can be used as a modifier as follows:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Specifying the default modifier}
\NormalTok{default_modifiers <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{branch.length =}\NormalTok{ branch.length)}

\CommentTok{## Setting some parameters for generating trees}
\NormalTok{bd_params  <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{extinction =} \DecValTok{0}\NormalTok{)}
\NormalTok{stope_rule <-}\StringTok{ }\KeywordTok{list}\NormalTok{(}\DataTypeTok{max.living =} \DecValTok{20}\NormalTok{)}

\CommentTok{## Generating a tree with the default branch length parameter}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
\NormalTok{default_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
                     \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
                     \DataTypeTok{modifiers =}\NormalTok{ default_modifiers)}
\end{Highlighting}
\end{Shaded}

Of course, the point of the modularity here is that you can provide your own function for generating branch length.
For example, we might be interested in what our tree would look like if we'd use a simple constant branch length generation (instead of randomly drawing it from an exponential distribution).
We can do so by declaring our own \texttt{branch.length} function and adding it to a \texttt{"modifiers"} object.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## A constant branch length generator}
\CommentTok{## (note that the output must be numeric, not integer)}
\NormalTok{constant.brlen <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{() \{}
    \KeywordTok{return}\NormalTok{(}\KeywordTok{as.numeric}\NormalTok{(}\DecValTok{1}\NormalTok{))}
\NormalTok{\}}

\CommentTok{## Creating the modifiers object}
\NormalTok{constant_modifier <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{branch.length =}\NormalTok{ constant.brlen)}

\CommentTok{## Generating a new tree with this modifier}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
\NormalTok{modified_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
                      \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
                      \DataTypeTok{modifiers =}\NormalTok{ constant_modifier)}
\end{Highlighting}
\end{Shaded}

And we can visualise the difference between both resulting trees:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(default_tree,  }\DataTypeTok{main =} \StringTok{"Default modifier"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(modified_tree, }\DataTypeTok{main =} \StringTok{"Constant branch length}\CharTok{\textbackslash{}n}\StringTok{modifier"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-45-1.pdf}

It is of course to use more complex branch length modifiers that intakes different conditions and specific modification rather than simply always output a value of one.

\hypertarget{allowarguments}{%
\subsection{The allowed arguments}\label{allowarguments}}

It is of course possible to design some more advanced function to interact with the birth death process.
You can create a function for \texttt{branch.length}, \texttt{selection} and \texttt{speciation} that involve any of the following arguments:

\begin{itemize}
\tightlist
\item
  \texttt{bd.params}: a named list containing \texttt{"numeric"} values that contains the birth death parameters (at least \texttt{"speciation"} and \texttt{"extinction"});
\item
  \texttt{lineage}: a named list containing the lineage data (see below).
\item
  \texttt{trait.values}: a \texttt{"matrix"} containing \texttt{"numeric"} values with the trait names as column names and the lineages ID as row numbers (you can use it with the function \texttt{parent.traits} to access the trait of the previous node for example).
\item
  \texttt{modify.fun}: a \texttt{"list"} of named \texttt{"function"} (usually passed through \texttt{condition} and \texttt{modify}).
\end{itemize}

The \texttt{lineage} list contains the following elements (missing elements are allowed):

\begin{itemize}
\tightlist
\item
  \texttt{lineage\$parents}: an \texttt{"integer"} vector: the list of parent lineages;
\item
  \texttt{lineage\$livings}: an \texttt{"integer"} vector: the list of lineages still not extinct;
\item
  \texttt{lineage\$drawn}: a single \texttt{"integer"}: the ID of the selected lineage;
\item
  \texttt{lineage\$current}: a single \texttt{"integer"}: the selected lineage (is equal to \texttt{lineage\$livings{[}lineage\$drawn{]}});
\item
  \texttt{lineage\$n}: a single \texttt{"integer"}: the current number of non extinct lineage (is equal to \texttt{length(lineage\$livings))};
\item
  \texttt{lineage\$split}: a \texttt{"logical"} vector: the list of splits for each lineage (\texttt{TRUE}), the number of total tips is equal to \texttt{sum(!lineage\$split)}.
\end{itemize}

In general, unless you know what you're doing, you can ignore most arguments for specific modifiers since they are handled automatically within the \texttt{dads} function.
Therefore any argument can be left undeclared or missing and is always handled internally. For example, if you did not declare \texttt{n.taxa} as a function argument but are using \texttt{n.taxa} in the function, \texttt{n.taxa} will be detected and treated as a current argument automatically as set accordingly within the birth death process (e.g. \texttt{n.taxa} will be set to the current number of taxa every iteration of the process).

For example, we can create a function that increases branch length proportional to the number of species ``alive'' at each time of the simulation in a discrete way. I.e. for discrete numbers of taxa, the branch length increases by jumps (ten fold):

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## A more complex binned.branch.length function}
\NormalTok{increasing.brlen <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(bd.params, lineage) \{}

    \CommentTok{## Setting the cumulated birth and death}
\NormalTok{    birth_death <-}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{speciation }\OperatorTok{+}\StringTok{ }\NormalTok{bd.params}\OperatorTok{$}\NormalTok{extinction}

    \CommentTok{## Returning branch lengths depending on different number of taxa}
    \ControlFlowTok{if}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{<=}\StringTok{ }\DecValTok{5}\NormalTok{) \{}
        \KeywordTok{return}\NormalTok{(}\DecValTok{1}    \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(}\DecValTok{5} \OperatorTok{*}\StringTok{ }\NormalTok{birth_death)))}
\NormalTok{    \}}
    \ControlFlowTok{if}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{<=}\StringTok{ }\DecValTok{10}\NormalTok{) \{}
        \KeywordTok{return}\NormalTok{(}\DecValTok{10}   \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(}\DecValTok{10} \OperatorTok{*}\StringTok{ }\NormalTok{birth_death)))   }
\NormalTok{    \}}
    \ControlFlowTok{if}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{<=}\StringTok{ }\DecValTok{15}\NormalTok{) \{}
        \KeywordTok{return}\NormalTok{(}\DecValTok{100}  \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(}\DecValTok{15} \OperatorTok{*}\StringTok{ }\NormalTok{birth_death)))      }
\NormalTok{    \}}
    \ControlFlowTok{if}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{<=}\StringTok{ }\DecValTok{20}\NormalTok{) \{}
        \KeywordTok{return}\NormalTok{(}\DecValTok{1000} \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(}\DecValTok{20} \OperatorTok{*}\StringTok{ }\NormalTok{birth_death)))}
\NormalTok{    \} }\ControlFlowTok{else}\NormalTok{ \{}
        \KeywordTok{return}\NormalTok{(}\DecValTok{1000} \OperatorTok{*}\StringTok{ }\KeywordTok{rexp}\NormalTok{(}\DecValTok{1}\NormalTok{, }\KeywordTok{sum}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n }\OperatorTok{*}\StringTok{ }\NormalTok{birth_death)))}
\NormalTok{    \}}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

We can then create it as a \texttt{"modifiers"} object and run a new simulation:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Creating a modifiers}
\NormalTok{increasing_modifier <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{branch.length =}\NormalTok{ increasing.brlen)}

\CommentTok{## Generating a new tree with this modifier}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
\NormalTok{increasing_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
                        \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
                        \DataTypeTok{modifiers =}\NormalTok{ increasing_modifier)}
\end{Highlighting}
\end{Shaded}

And we can visualise the difference between the resulting trees:

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{3}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(default_tree,    }\DataTypeTok{main =} \StringTok{"Default modifier"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(modified_tree,   }\DataTypeTok{main =} \StringTok{"Constant branch length}\CharTok{\textbackslash{}n}\StringTok{modifier"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(increasing_tree, }\DataTypeTok{main =} \StringTok{"Increasing branch length}\CharTok{\textbackslash{}n}\StringTok{modifier (binned)"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-48-1.pdf}

\hypertarget{the-selection-function-selection}{%
\section{\texorpdfstring{The selection function (\texttt{selection})}{The selection function (selection)}}\label{the-selection-function-selection}}

The \texttt{selection} function is used in the birth death process to know which lineage to select when running a speciation (or extinction!) event.
By default, this function randomly selects one taxon that is currently not extinct (using: \texttt{sample(1:n.taxa,\ 1))}.
Similarly as \texttt{branch.length} it is possible to modify this part of the birth death process.
For example, we could simply select always the last created lineage (to create a ``ladder'' or most asymmetric tree):

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Our function to always select the last taxon}
\CommentTok{## (making sure it returns an integer)}
\NormalTok{select.last <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(lineage) \{}
    \KeywordTok{return}\NormalTok{(}\KeywordTok{as.integer}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{n))}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

\begin{quote}
Note that here the function can only intake the \protect\hyperlink{allowarguments}{allowed arguments as described above} (here \texttt{n.taxa}: the number of current living taxa).
\end{quote}

We can then create a \texttt{"modifiers"} object the same way as before using this time the \texttt{selection} argument:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## A modifier for selection}
\NormalTok{ladderised_modifier <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{selection =}\NormalTok{ select.last)}
\CommentTok{## Generating a new tree with this modifier}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{0}\NormalTok{)}
\NormalTok{ladderised_tree <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
                        \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
                        \DataTypeTok{modifiers =}\NormalTok{ ladderised_modifier)}
\CommentTok{## Displaying the results}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,}\DecValTok{2}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(default_tree,    }\DataTypeTok{main =} \StringTok{"Default modifier"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(ladderised_tree, }\DataTypeTok{main =} \StringTok{"Ladderising modifier"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-50-1.pdf}

Again, it is of course possible to make the modifier more complex and in combination with other elements of the tree.
For example, we can create a \texttt{"dads"} object that also creates a trait a add to it a \texttt{selection} modifier that only selects for tips with positive trait values.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Our function that only select taxa with positive trait values}
\NormalTok{select.positive <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{(trait.values, lineage) \{}
    \CommentTok{## Selecting the taxa with positive traits only}
\NormalTok{    positives <-}\StringTok{ }\KeywordTok{as.integer}\NormalTok{(}\KeywordTok{rownames}\NormalTok{(trait.values)[}\KeywordTok{which}\NormalTok{(trait.values[, }\DecValTok{1}\NormalTok{] }\OperatorTok{>=}\StringTok{ }\DecValTok{0}\NormalTok{)])}

    \CommentTok{## Select the current taxa that descend from a node with a positive value}
\NormalTok{    positive_living <-}\StringTok{ }\KeywordTok{cbind}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{parents, }\KeywordTok{seq_along}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{split)}
\NormalTok{                             )[}\KeywordTok{which}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{parents }\OperatorTok{%in%}\StringTok{ }\NormalTok{positives), }\DecValTok{2}\NormalTok{]}

    \CommentTok{## Select one tip randomly in the ones with descendants with positive values}
    \KeywordTok{return}\NormalTok{(}\KeywordTok{sample}\NormalTok{(}\KeywordTok{which}\NormalTok{(lineage}\OperatorTok{$}\NormalTok{livings }\OperatorTok{%in%}\StringTok{ }\NormalTok{positive_living), }\DecValTok{1}\NormalTok{))}
\NormalTok{\}}

\CommentTok{## Creating the modifier}
\NormalTok{positive_skew <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{selection =}\NormalTok{ select.positive)}

\CommentTok{## Creating a (default) trait object}
\NormalTok{BM_trait <-}\StringTok{ }\KeywordTok{make.traits}\NormalTok{()}

\CommentTok{## Simulate a tree and trait with no modifier}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{default_dads <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
                     \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
                     \DataTypeTok{traits    =}\NormalTok{ BM_trait)}

\CommentTok{## Simulate a tree and trait with the modifier}
\KeywordTok{set.seed}\NormalTok{(}\DecValTok{1}\NormalTok{)}
\NormalTok{skewed_trait_dads <-}\StringTok{ }\KeywordTok{dads}\NormalTok{(}\DataTypeTok{bd.params =}\NormalTok{ bd_params,}
                          \DataTypeTok{stop.rule =}\NormalTok{ stope_rule,}
                          \DataTypeTok{traits    =}\NormalTok{ BM_trait,}
                          \DataTypeTok{modifiers =}\NormalTok{ positive_skew)}

\CommentTok{## Plotting the differences in trees and traits}
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mfrow =} \KeywordTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\DecValTok{2}\NormalTok{))}
\KeywordTok{plot}\NormalTok{(default_dads, }\DataTypeTok{main =} \StringTok{"Default trait and tree"}\NormalTok{)}
\KeywordTok{plot}\NormalTok{(skewed_trait_dads, }\DataTypeTok{main =} \StringTok{"Skewed trait and tree"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{dads_manual_files/figure-latex/unnamed-chunk-51-1.pdf}

@@

The third function that can be used to modify the birth death process is the \texttt{speciation} function.
This one is used during the birth death process to decide whether a lineages speciates (creating a node and two new lineages) or goes extinct (creating a tip).

\begin{quote}
Note that the \texttt{speciation} function only affects tips or nodes before the simulation reaches the \texttt{stop.rule}.
The then surviving lineages are all automatically transformed into tips.
\end{quote}

By default, the \texttt{speciation} function is trigger a speciation even if a number randomly drawn from a uniform distribution is lower than the ratio between the speciation and the speciation and extinction parameter. If the randomly drawn number is higher, the lineage goes extinct.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## The speciation in pseudo-code:}
\KeywordTok{runif}\NormalTok{(}\DecValTok{1}\NormalTok{) }\OperatorTok{<}\StringTok{ }\NormalTok{speciation}\OperatorTok{/}\StringTok{ }\NormalTok{(speciation }\OperatorTok{+}\StringTok{ }\NormalTok{extinction)}
\end{Highlighting}
\end{Shaded}

Creating a \texttt{"modifiers"} with a \texttt{speciation} function works the same way as for \texttt{branch.length} and \texttt{selection} but the function that will be used needs to output a logical value (see table \protect\hyperlink{summarymodifiers}{below}).
Once the function is created simply input your function for speciation in the modifier and run the \texttt{dads} function with that modifier:

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{## Speciating or going extinct randomly}
\CommentTok{## (regardless of the extinction parameter)}
\NormalTok{random.extinct  <-}\StringTok{ }\ControlFlowTok{function}\NormalTok{() \{}
    \KeywordTok{return}\NormalTok{(}\KeywordTok{sample}\NormalTok{(}\KeywordTok{c}\NormalTok{(}\OtherTok{TRUE}\NormalTok{, }\OtherTok{FALSE}\NormalTok{), }\DecValTok{1}\NormalTok{))}
\NormalTok{\}}

\CommentTok{## Creating the modifiers object}
\NormalTok{random_extinction <-}\StringTok{ }\KeywordTok{make.modifiers}\NormalTok{(}\DataTypeTok{speciation =}\NormalTok{ random.extinct)}
